\pkg{NeuralNetTools}: Visualization and Analysis Tools for Neural Networks

### 2016-11-22 nnt_manu.tex
revision draft done, need to proof and fix applied example

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  Pruning has the additional benefit of increasing power by increasing the degrees of freedom or the identification of input nodes with low importance.  Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

@@

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).  All explanatory variables were normalized and the response variable was standardized to 0--1.

The effects of network architecture and initial starting weights on estimates of variable importance were evaluated by creating many models and assessing the uncertainty of the importance.  Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

Several conclusions can be made from \cref{fig:flightimp}.  First, delays in arrival time are negatively related to distance and positively related to departure delays and air time.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
\label{fig:flightimp1}
}

>>>>

The \pkg{RSNNS} package provides algorithms to prune connections or nodes in a neural network \citep{Bergmeir12}.  This approach can remove connection weights between layers or input nodes that do not contribute to the predictive performance of the network.  In addition to visualizing connections in the network that are not important, connections that are pruned can be removed in successive model fitting.  This reduces the number of free parameters (weights) that are  estimated by the model optimization algorithm, thereby increasing the likelihood of convergence to an estimable numeric solution for the remaining connection weights that minimizes prediction error \citep[i.e., model identifiability,][]{Ellenius00}. Algorithms in \pkg{RSNNS} for weight pruning include magnitude-based pruning, Optimal Brain Damage, and Optimal Brain Surgeon, whereas algorithms for node pruning include skeletonization and the non-contributing units method \citep{Zell98}.  The \code{plotnet} function can plot a pruned neural network, with options to omit or display the pruned connections (\cref{fig:plotprune}).

@@

A dataset from \pkg{nycflights13} \citep{Wickham14b} is used to demonstrate 1) use of the functions in \pkg{NeuralNetTools} to gain additional insight into relationships among variables, and 2) effects of training conditions on model conclusions (get analysis code \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{here}).  This dataset provides information on all flights departing New York City (i.e., JFK, LGA, or EWR) in 2013.  The example uses all flights from the UA carrier in the month of December to identify variables that potentially influence arrival delays (\code{arr\_delay}, minutes) at the destination airport.  Factors potentially related to delays were selected from the dataset and included departure delay (\code{dep\_delay}, minutes), departure time (\code{dep\_time}, hours, minutes), arrival time (\code{arr\_time}, hours, minutes), travel time between destinations (\code{air\_time}, minutes), and distance flown (\code{distance}, miles).

First, the apropriate month and airline carrier were selected, all explanatory variables were normalized, and the response variable was standardized to 0--1.
\begin{Schunk}
\begin{Sinput}
R> library("nycflights13")
R> library("dplyr")
R> tomod <- filter(flights, month == 12 & carrier == 'UA') %>% 
+   select(arr_delay, dep_delay, dep_time, arr_time, air_time, distance) %>%
+   mutate_each(funs(scale), -arr_delay) %>% 
+   mutate_each(funs(as.numeric), -arr_delay) %>% 
+   mutate(arr_delay = scales::rescale(arr_delay, to = c(0, 1))) %>% 
+   data.frame
\end{Sinput}
\end{Schunk}
Then, a standard \ac{mlp} with five hidden nodes was created with the \pkg{nnet} package to model the effects of selected variables on arrival delays. The entire dataset is used for the example but separate training and validation datasets should be used in practice.  
\begin{Schunk}
\begin{Sinput}
R> library(nnet)
R> mod <- nnet(arr_delay ~ ., size = 5, linout = TRUE, data = tomod, trace = F)  
\end{Sinput}
\end{Schunk}
The default output is limited to structural information about the model and methods for model predictions (see \code{str(mod)} and \code{?predict.nnet}). Using functions from \pkg{NeuralNetTools}, a more comprehensive understanding of the relationships between the variables is illustrated.
\begin{Schunk}
\begin{Sinput}
R> plotnet(mod)
R> garson(mod)
R> olden(mod)
R> lekprofile(mod, group_vals = 5)
R> lekprofile(mod, group_vals = 5, group_show = TRUE)
\end{Sinput}
\end{Schunk}

\Cref{fig:flightplots} shows the information about arrival delays that can be obtained with the functions in \pkg{NeuralNetTools}. The \ac{nid} (\ref{fig:flightplotsa}) shows the model structure and can be used develop a general characterization of the relationships between variables.  For example, most of the connection weights from input node I5 are negative (grey), suggesting that distance travelled has an opposing relationship with arrival delays. \Cref{fig:flightplotsb} and \ref{fig:flightplotsc} provide more quantitative descriptions using information from both the \ac{nid} and model predictions.  \Cref{fig:flightplotsb} shows variable importance using the \code{garson} and \code{olden} algorithms.  The \code{garson} function suggests time between destinations (\code{air\_time}) has the strongest relationship with arrival delays, similar to a strong positive association shown with the \code{olden} method.  However, the \code{garson} function shows arrival time as the second most important variable, whereas the \code{olden} function shows this variable as having a weak negative relationship with arrival time. This discrepancy is explained below. Finally, results from the \code{lekprofile} function (\ref{fig:flightplotsc}) confirm those in \ref{fig:flightplotsb}, with the addition of non-linear responses that vary by different groupings of the data.  Values for each variable in the different unevaluated groups (based on clustering) show that there were no obvious patterns between groups with the exception being group four that generally had early arrival times and long departure delays.

%%%%
\begin{figure}
\centering
\subfloat[][\code{plotnet}]{
\includegraphics[width = 0.5\textwidth, page = 1]{figs/flightplots1.pdf}
\label{fig:flightplotsa}
}
\subfloat[][\code{garson} (top) and \code{olden} (bottom)]{
\includegraphics[width = 0.5\textwidth, page = 2]{figs/flightplots1.pdf}
\label{fig:flightplotsb}
}

\subfloat[][\code{lekprofile}]{
\includegraphics[width = 1.05\textwidth, page = 1]{figs/flightplots2.pdf}
\label{fig:flightplotsc}
}

\caption{Results from a simple \ac{mlp} model of arrival delay for December airline flights versus departure delay (\code{dep\_delay}), departure time (\code{dep\_time}), arrival time (\code{arr\_time}), travel time between destinations (\code{air\_time}), and distance flown (\code{distance}). The three plots show the \ac{nid} from \code{plotnet} (\ref{fig:flightplotsa}), variable importance with \code{garson} and \code{olden} (\ref{fig:flightplotsb}), and sensitivity analysis with variable groupings from \code{lekprofile} (\ref{fig:flightplotsc}). Interpretations are provided in the text.}
\label{fig:flightplots}
\end{figure}

A second analysis was needed to show the effects of network architecture and initial starting weights on uncertainty in estimates of variable importance. Models with one, five, or ten hidden nodes and 100 separate models for each node level were created.  Each model had a random set of starting weights for the first training iteration.  Importance estimates using \code{olden} were saved for each model and combined in a single plot to show overall variable importance as the median and 5\textsuperscript{th}/95\textsuperscript{th} percentiles from the 100 models for each node level (see analysis \href{https://raw.githubusercontent.com/fawda123/nnt_manu/master/flightimp.R}{code}).

Several conclusions from \cref{fig:flightimp} provide further information to interpret the trends in \cref{fig:flightplots}.  First, consistent relationships can be identified such that delays in arrival time were negatively related to distance and positively related to departure delays and air time.  That is, flights arrived later than their scheduled time if flight time was long or if their departure was delayed, whereas flights arrived earlier than scheduled for longer distances.  No conclusions can be made for the other variables because the bounds of uncertainty included zero.  Second, the range of importance estimates varied between the models (i.e., one node  varied +/- 1 and the others +/- 3).  This suggests that the relative importance estimates only have relevance within each model, whereas only the rankings (e.g., least, most important) can be compared between models.  Third and most important, the level of uncertainty for specific variables can be large between model fits for the same architecture.  This suggests that a single model can provide misleading information and therefore several models may be required to decrease uncertainty.  Additional considerations described above (e.g., criteria for stopping training, use of training and test datasets) can also affect the interpration of model information and should be considered equally during model development.
\begin{figure}[!h]
\centering
\subfloat[][Networks with one node]{
\includegraphics[width = 0.95\textwidth, page = 1]{figs/flightimp.pdf}
\label{fig:flightimp1}
}
